---
title: "Bayesian Multilevel Models in R with `brms`"
output:
  html_notebook:
    toc: yes
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = ">#")
```

In this note I'll replicate the models I demonstrated in the workshop with
Bayesian estimation and the `brms` package. The `brms` package depends on
installation of
[`rstan`](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started), which
currently does not work on RStudio Cloud. If you're using your own computer,
please go to https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started to
have `rstan` installed first, and then install `brms`.

# 1. Importing Data Into R

We'll import the data in .sav format. 

```{r import_sav, message=FALSE}
# install.packages("tidyverse")  # if you haven't installed the `tidyverse` package
library(tidyverse)
theme_set(theme_bw())  # Use a different theme for plotting with ggplot2
library(haven)
# Read in the data (pay attention to the directory)
hsball <- read_sav("../data/hsball.sav")
hsball  # print the first 10 rows
```

# 2. Random Intercept Model

First load the package:

```{r load_lme4, message=FALSE}
# install.packages("brms")  # if you haven't installed the `brms` package, which
# requires `rstan` to be installed first
# Load the `brms` library to run MLM
library(brms)
options(mc.cores = 2L)  # use two cores for parallel processing
```

```{r ran_int, results='hide'}
# Simply change `lmer()` to `brm()` (which uses the default non-informative
# priors); I change the # of chains and iterations for shorter run time. 
# It takes about a minute for compilation, and then about 
ran_int <- brm(mathach ~ (1 | id), data = hsball, chains = 2, iter = 1000)
```

You can summarize the results using

```{r summary_ran_int}
summary(ran_int)
```

# 3. Model With a Lv-2 predictor

```{r m_lv2, results='hide'}
m_lv2 <- brm(mathach ~ meanses + (1 | id), data = hsball, chains = 2, 
             iter = 1000)
```

```{r summary_m_lv2}
summary(m_lv2)
```

# 4. Examing Contextual Effects

```{r contextual, results='hide'}
contextual <- brm(mathach ~ meanses + ses + (1 | id), data = hsball, 
                  chains = 2, iter = 1000)
```

```{r summary_contextual}
summary(contextual)
```

# 5. Run the Random-Coefficients Model

```{r ran_slp, results='hide'}
ran_slp <- brm(mathach ~ meanses + ses + (ses | id), data = hsball, 
                  chains = 2, iter = 1000)
```

```{r summary_ran_slp}
summary(ran_slp)
```

# 6. Analyzing Cross-Level Interactions

```{r crlv_int, results='hide'}
crlv_int <- brm(mathach ~ meanses + sector * ses + (ses | id), 
                 data = hsball, chains = 2, iter = 1000)
```

```{r summary_crlv_int}
summary(crlv_int)
```
