---
title: "Multilevel Modeling Diagnostics in R"
output:
  html_notebook:
    toc: yes
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = ">#")
```

In R, there are many quite a few diagnostic tools for MLM, although they are 
not necessarily straightforward to use. In this note I will show some basic 
tools. 

# Running MLM

## Importing Data Into R

We'll import the data in .sav format. 

```{r import_sav}
library(tidyverse)
theme_set(theme_bw())  # Use a different theme for plotting with ggplot2
library(lme4)
library(haven)
# Read in the data (pay attention to the directory)
hsball <- read_sav("../data/hsball.sav")
hsball <- as.data.frame(hsball)  # need this to get the influence statistics
```

## Fitting the model

We will use the example of a cross-level interaction model

```{r crlv_int}
crlv_int <- lmer(mathach ~ meanses + sector * ses + (ses | id), 
                 data = hsball)
```

# Diagnostic Plots

## 1. Residuals (Lv 1.) vs. Fitted Values

For: linearity, specification, outliers, homoscedasticity

You want to see a random blob with no clear patterns, such as a curvillinear
shape

```{r resid_vs_fitted}
# Augment the data with the fitted values and the standardized residuals
resid_df <- mutate(hsball, 
                   .std_resid = resid(crlv_int, scaled = TRUE), 
                   .fitted = fitted(crlv_int))
ggplot(resid_df, 
       aes(x = .fitted, y = .std_resid)) + 
  geom_point() +  # show the points
  geom_smooth()  # add a smoother
```

The plot shows no clear patterns and there does not appear to be outliers. 
The residual variance also seems preety constant across different levels of 
the fitted values. *One thing to note, however, is that the residuals appear to
be bounded, which can indicate a ceiling and/or a floor effect. This certainly
is due to our outcome variable being bounded with a maximum of 25. 

## 2. Residuals (Lv 1.) vs. Each Predictor

For: linearity, specification, outliers, homoscedasticity

You want to see a random blob with no clear patterns, such as a curvillinear
shape

### a. For `ses`

```{r resid_vs_ses}
p1 <- ggplot(resid_df, 
       aes(x = ses, y = .std_resid)) + 
  geom_point() +  # show the points
  geom_smooth()  # add a smoother
```

### b. For `meanses`

```{r resid_vs_meanses}
p2 <- ggplot(resid_df, 
       aes(x = meanses, y = .std_resid)) + 
  geom_point() +  # show the points
  geom_smooth()  # add a smoother
```

### c. For `sector`

For categorical predictors, you want to see that the variance and distribution
is similar across different levels of the predictor. 

```{r resid_vs_sector}
p3 <- ggplot(resid_df, 
       aes(x = factor(sector),  # factor() is used for categorical predictor
           y = .std_resid)) + 
  # use `geom_violin()` with categorical predictor
  geom_violin()
gridExtra::grid.arrange(p1, p2, p3, nrow = 2L)
```

The plots did not suggest any major issues with the linearity or homogeneity. 

## Q-Q Plot for Lv-1 Residuals

For: normality

You want to see the points line up nicely along the 45 degree line

```{r qq_lv1}
library(lattice)  # need this package to use the built-in functions
qqmath(crlv_int)  # just use the `qqmath()` function on the fitted model
```

There are some deviations from normality on the two tail areas, which is likely
due to less kurtosis because of the boundedness of the outcome variable. 
Generally the impact of kurtosis is not as severe as skewness, and the above 
plot did not suggest a major issue to the results. 

## Q-Q Plot for Lv-2 Residuals

For: normality

At level 2 there can be more then one set of residuals. For example, in this 
model, we have one set for the random intercept, and the other set for the 
random slope

```{r qq_lv2}
# the code is a bit more complex in order to add, but you only need to replace
# the reference lines `crlv_int` with the name of your own model to get the
# plots for your own analyses
qqmath(ranef(crlv_int), 
       panel = function(x) {  
         panel.qqmath(x)
         panel.qqmathline(x)
       })
```

Normality does not appear to be an issue at level 2. 

## Variance Inflation Factor (VIF)

For: multicollinearity

```{r vif_model}
library(car)
car::vif(crlv_int)
```

As a rule of thumb, one would worry about multicollinearity in the predictors 
when VIF > 4, and VIF > 10 indicates strong multicollinearity. 

## Influence Index Plot

For: leverage and outliers

To examine whether some cases gives disproportionately strong influence on the
results, we can look at influence index plot and dfbeta

```{r influence_plots, fig.height=10, fig.width=6}
inf_crlv_int <- influence(crlv_int, "id")  # get influence statistics
car::infIndexPlot(inf_crlv_int)
```

If you need some cutoffs, the cutoff for dfbeta is $2 / sqrt{n}$, which would
be `r 2 / sqrt(160)` in our example; cutoff for Cook's distance is 
$4 / (n - 2)$, which would be `r 4 / (160 - 2)` in our example. 
